---
title: "Global Health Insights Explorer"
subtitle: "MVP aligned with proposal"
author: "Vera Zhang"
format: html
execute:
  warning: false
  message: false
---

## Project overview

This minimum viable product (MVP) retrieves **global health indicator data directly from the World Bank API**, cleans it in R, visualizes trends across countries, and produces a short plain-language interpretation. It is a first working version of the product proposed in the course proposal.

## What this page does

- Pulls data from the **World Bank Open Data API**
- Compares **life expectancy at birth** across selected countries
- Displays an interactive plot
- Generates a concise interpretation based on the most recent available values

```{r}
#| label: setup
#| include: false

library(httr2)
library(jsonlite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(glue)
library(scales)

country_codes <- c("US", "CN", "BR")
country_names <- c(
  US = "United States",
  CN = "China",
  BR = "Brazil"
)

indicator_code <- "SP.DYN.LE00.IN"
indicator_name <- "Life expectancy at birth, total (years)"
start_year <- 2000
end_year <- 2023

fetch_world_bank_indicator <- function(countries, indicator, start_year = 2000, end_year = 2023) {
  url <- glue(
    "https://api.worldbank.org/v2/country/{paste(countries, collapse = ';')}/indicator/{indicator}?format=json&per_page=20000"
  )

  resp <- request(url) |>
    req_perform()

  txt <- resp_body_string(resp)
  parsed <- fromJSON(txt, flatten = TRUE)

  if (length(parsed) < 2 || is.null(parsed[[2]])) {
    stop("No data returned from the World Bank API.")
  }

  dat <- as_tibble(parsed[[2]]) |>
    transmute(
      country_code = countryiso3code,
      country = country.value,
      year = as.integer(date),
      value = as.numeric(value)
    ) |>
    filter(!is.na(value), year >= start_year, year <= end_year)

  dat
}

wb_raw <- fetch_world_bank_indicator(country_codes, indicator_code, start_year, end_year)

wb_data <- wb_raw |>
  mutate(
    country = recode(country_code,
      USA = "United States",
      CHN = "China",
      BRA = "Brazil",
      .default = country
    )
  ) |>
  arrange(country, year)

latest_data <- wb_data |>
  group_by(country) |>
  filter(year == max(year, na.rm = TRUE)) |>
  ungroup() |>
  arrange(desc(value))

summary_tbl <- latest_data |>
  select(country, year, value) |>
  rename(`Most recent year` = year, `Life expectancy` = value)
```

## Data preview

```{r}
summary_tbl
```

## Interactive trend plot

```{r}
p <- ggplot(wb_data, aes(x = year, y = value, color = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  labs(
    title = "Life expectancy trends",
    subtitle = "World Bank API data",
    x = "Year",
    y = "Life expectancy at birth (years)",
    color = "Country"
  ) +
  theme_minimal(base_size = 12)

p
```

## Automatically generated interpretation

```{r}
top_country <- latest_data$country[1]
top_value <- latest_data$value[1]
top_year <- latest_data$year[1]

bottom_country <- latest_data$country[nrow(latest_data)]
bottom_value <- latest_data$value[nrow(latest_data)]

change_tbl <- wb_data |>
  group_by(country) |>
  summarise(
    first_year = min(year, na.rm = TRUE),
    last_year = max(year, na.rm = TRUE),
    first_value = value[year == first_year][1],
    last_value = value[year == last_year][1],
    change = last_value - first_value,
    .groups = "drop"
  ) |>
  arrange(desc(change))

largest_gain_country <- change_tbl$country[1]
largest_gain_value <- change_tbl$change[1]

cat(glue(
  'Using the most recent available World Bank data, **{top_country}** had the highest life expectancy among the selected countries at **{round(top_value, 1)} years** in **{top_year}**, while **{bottom_country}** had the lowest at **{round(bottom_value, 1)} years**. From **{start_year}** to **{end_year}**, **{largest_gain_country}** showed the largest overall increase, with life expectancy rising by approximately **{round(largest_gain_value, 1)} years**. This MVP demonstrates the core pipeline proposed for the final project: API retrieval, data cleaning, visualization, and automated interpretation.'
))
```

## Notes for next iteration

This MVP is intentionally simple so that the pipeline works end to end. The next version can extend the proposal by adding:

1. User-selected countries and indicators
2. Additional World Bank health indicators
3. A true GenAI summary step using Gemini or OpenAI API
4. More comparative views and richer analytical components

## Reproducibility

```{r}
sessionInfo()
```